---
title: "Tutorial: Creating Spatial Data for NEON Plot Subtypes"
author: "Your Name"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Creating Spatial Data for NEON Plot Subtypes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(sf)
library(lwgeom)
library(ggspatial)
source(paste0(getwd(), "/R/config.R"))
```

# Introduction

This tutorial demonstrates how to generate and visualize spatial data for different NEON (National Ecological Observatory Network) plot subtypes using R. We'll create basic maps to visualize spatial relationships and calculate relevant attributes, like plot radii.

# Load Required Packages and Set Configurations

We begin by loading required libraries and setting the projection for spatial data.

```{r load-packages}
# Specify projection
prj <- "EPSG:5070" # Albers Equal Area projection
```

# Functions for Spatial Calculations

Two utility functions are used:
- `id_circle_center()`: Identifies the centroid of the minimum bounding circle for a set of geometries.
- `max_dist_to_circle_center()`: Calculates the maximum distance from the centroid to the geometries' boundary.

```{r define-functions}
id_circle_center <- function(d) {
  du <- st_union(d)
  b <- lwgeom::st_minimum_bounding_circle(du)
  c <- st_centroid(b)
  return(c)
}

max_dist_to_circle_center <- function(d) {
  du <- st_union(d)
  b <- lwgeom::st_minimum_bounding_circle(du)
  c <- st_centroid(b)
  dp <- d %>% st_cast("POINT")
  dist <- st_distance(dp, c)
  return(as.numeric(max(dist)))
}
```

# Load and Clean Data

Next, we import and preprocess NEON spatial data, ensuring consistent projections and joining geometries.

```{r load-data}
# Load NEON domains
dom <- st_read("data/NEON_Domains.shp", quiet = TRUE) %>%
  st_transform(prj) %>%
  rename(domainNumb = DomainID, domainName = DomainName) %>%
  mutate(domainNumb = factor(sprintf("D%02d", domainNumb))) %>%
  group_by(domainNumb) %>%
  summarize(domainName = first(domainName), geometry = st_union(geometry))

# Load NEON sites
site <- st_read("data/terrestrialSamplingBoundaries.shp", quiet = TRUE) %>%
  st_transform(prj) %>%
  group_by(siteID) %>%
  summarize(domainName = first(domainName), domainNumb = first(domainNumb), geometry = st_union(geometry))
```

# Visualize NEON Domains

We visualize NEON domains to understand their spatial extent.

```{r plot-domains}
ggplot() +
  geom_sf(data = dom, fill = "lightblue", color = "darkblue") +
  ggtitle("NEON Domains") +
  theme_minimal()
```

# Filtering NEON Plots by Subtype

For this tutorial, we focus on the `mammalGrid` subtype.

```{r filter-plots}
plt <- st_read("data/NEON_TOS_Plot_Points.shp", quiet = TRUE) %>%
  filter(subtype == "mammalGrid") %>%
  st_transform(prj) %>%
  left_join(., st_drop_geometry(site %>% select(siteID, domainName, domainNumb)), by = c("siteID"))
```

# Calculating Plot-Level Buffers

We calculate plot-level centroids and radii, then buffer them to create circular polygons.

```{r calculate-plot-buffers}
plt_nested <- plt %>%
  nest(.by = c(plotID, plotType, siteID, domainName, domainNumb)) %>%
  mutate(
    circle_center = lapply(data, function(x) id_circle_center(x)),
    max_dist_circ = lapply(data, function(x) max_dist_to_circle_center(x))
  ) %>%
  unnest(cols = c(max_dist_circ, circle_center)) %>%
  mutate(plot_poly = st_buffer(circle_center, dist = 100))
```

# Visualize Plot Buffers

We map the plot polygons to verify their spatial layout.

```{r plot-buffers}
ggplot() +
  geom_sf(data = plt_nested, aes(fill = siteID), color = "black") +
  ggtitle("Plot Buffers") +
  theme_minimal()
```

# Calculating Site-Level Buffers

Similarly, site-level centroids and radii are calculated, buffering to 15 km.

```{r calculate-site-buffers}
site_nested <- plt %>%
  nest(.by = c(siteID, domainName, domainNumb)) %>%
  mutate(
    circle_center = lapply(data, function(x) id_circle_center(x)),
    max_dist_circ = lapply(data, function(x) max_dist_to_circle_center(x)),
    buff_dist = 15000
  ) %>%
  unnest(cols = c(max_dist_circ, circle_center)) %>%
  mutate(site_poly = st_buffer(circle_center, dist = buff_dist))
```

# Visualize Site Buffers

The site polygons are displayed to observe their spatial distribution.

```{r site-buffers-map}
ggplot() +
  geom_sf(data = site_nested, aes(fill = domainName), color = "darkgreen") +
  ggtitle("Site Buffers") +
  theme_minimal()
```

# Summary

This tutorial outlined:
1. Loading and cleaning NEON spatial data.
2. Filtering plots by subtype.
3. Calculating plot and site buffers.
4. Visualizing the spatial relationships.

By following these steps, you can create spatial data tailored to NEON plot subtypes and adapt this approach for other subtypes or spatial analyses.
```